---
title: "Baseball Analysis"
execute:
  echo: false
  warning: false
---

```{r}
#| message: false
library(tidyverse)
library(plotly)
# Statcast & pre-Statcast Hard-Hit Rate data (player-seasons)
hitters_hard_hit <- read_csv("data/all_hitters_hard_hit.csv")

yearly_avg_hh <- hitters_hard_hit |>
  group_by(year, era) |>
  summarize(avg_hard_hit = mean(hard_hit_pct, na.rm = TRUE), .groups = "drop")

# Pitcher seasons data (K/9)
pitcher_seasons <- read_csv("data/pitcher_seasons.csv") |>
  left_join(
    Lahman::People |> select(playerID, nameFirst, nameLast) |> mutate(name = paste(nameFirst, nameLast)),
    by = "playerID"
  )

yearly_pitching <- pitcher_seasons |>
  group_by(yearID, era) |>
  summarize(avg_so9 = mean(so9), .groups = "drop")
```

The introduction of Major League Baseball's Statcast system in 2015 revolutionized how the sport is measured and played. By deploying high-resolution cameras and radar across every ballpark, Statcast enabled the tracking of granular metrics like exit velocity, launch angle, and pitch spin rates on every play.

This project examines how key baseball statistics evolved alongside this technological shift. By comparing trends across the pre-Statcast era (2005–2014) and post-Statcast era (2015–2025), we analyze whether the availability of tracking data coincided with noticeable changes in pitching strikeout rates (K/9) and hitter hard-hit percentages.

## Strikeout Rates Over Time

Each point represents an individual pitcher-season (minimum 30 IP), displaying real pitcher-level variation across thousands of observations. The solid line and points trace the annual league average K/9 across the pre-Statcast and post-Statcast eras.

```{r}
#| label: plot-k9-pitcher-seasons
p_k9 <- ggplot() +
  geom_jitter(
    data = pitcher_seasons,
    aes(
      x = yearID,
      y = so9,
      color = era,
      text = paste0(name, " (", yearID, ")<br>K/9: ", round(so9, 2))
    ),
    alpha = 0.15,
    width = 0.25,
    size = 1.2
  ) +
  geom_line(
    data = yearly_pitching,
    aes(x = yearID, y = avg_so9),
    color = "black",
    linewidth = 1.2
  ) +
  geom_point(
    data = yearly_pitching,
    aes(
      x = yearID,
      y = avg_so9,
      fill = era,
      text = paste0(yearID, " League Avg K/9: ", round(avg_so9, 2))
    ),
    shape = 21,
    size = 3,
    color = "black"
  ) +
  geom_vline(xintercept = 2014.5, linetype = "dashed", color = "darkred", linewidth = 1) +
  annotate(
    "text",
    x = 2014.8,
    y = 16,
    label = "Statcast Era Begins (2015)",
    color = "darkred",
    hjust = 0,
    size = 3.5
  ) +
  scale_x_continuous(breaks = seq(2005, 2025, by = 5)) +
  scale_color_manual(values = c("Pre-Statcast (2005–2014)" = "#377eb8", "Post-Statcast (2015–2025)" = "#e41a1c")) +
  scale_fill_manual(values = c("Pre-Statcast (2005–2014)" = "#377eb8", "Post-Statcast (2015–2025)" = "#e41a1c")) +
  labs(
    title = "Strikeouts per 9 Innings (K/9) by Pitcher-Season (2005–2025)",
    x = "Year",
    y = "Average K/9",
    color = "Era",
    fill = "Era"
  ) +
  theme_minimal()

ggplotly(p_k9, tooltip = "text", height = 420)
```

Between the pre-Statcast era (2005–2014) and the post-Statcast era (2015–2025), average pitcher strikeout rates increased from **7.21** to **8.63 K/9**, representing a **19.7% increase** across eras (and a **31.5% increase** from 6.55 in 2005 to 8.61 in 2025).

## Hard Hit Rate Over Time

Each point represents a qualifying hitter's individual hard-hit rate for that season, showing real individual-level variation across 4,251 player-seasons. The solid line and points trace the annual league average across the pre-Statcast era (BIS video-scouted Hard%) and the Statcast era (radar-measured exit velocity ≥ 95 mph).

```{r}
#| label: plot-hard-hit-player-seasons
p_hard_hit <- ggplot() +
  geom_jitter(
    data = hitters_hard_hit,
    aes(
      x = year,
      y = hard_hit_pct,
      color = era,
      text = paste0(name, " (", year, ")<br>Hard-Hit %: ", hard_hit_pct, "%")
    ),
    alpha = 0.2,
    width = 0.25,
    size = 1.2
  ) +
  geom_line(
    data = yearly_avg_hh,
    aes(x = year, y = avg_hard_hit, group = era),
    color = "black",
    linewidth = 1.2
  ) +
  geom_point(
    data = yearly_avg_hh,
    aes(
      x = year,
      y = avg_hard_hit,
      fill = era,
      text = paste0(year, " League Average: ", round(avg_hard_hit, 1), "%")
    ),
    shape = 21,
    size = 3,
    color = "black"
  ) +
  geom_vline(xintercept = 2014.5, linetype = "dashed", color = "darkred", linewidth = 1) +
  annotate(
    "text",
    x = 2014.8,
    y = 65,
    label = "Statcast Era Begins (2015)",
    color = "darkred",
    hjust = 0,
    size = 3.5
  ) +
  scale_x_continuous(breaks = seq(2005, 2025, by = 5)) +
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  scale_color_manual(values = c("Pre-Statcast (2005–2014)" = "#377eb8", "Post-Statcast (2015–2025)" = "#e41a1c")) +
  scale_fill_manual(values = c("Pre-Statcast (2005–2014)" = "#377eb8", "Post-Statcast (2015–2025)" = "#e41a1c")) +
  labs(
    title = "Hard-Hit Rate by Player-Season (2005–2025)",
    x = "Year",
    y = "Average Hard-Hit %",
    color = "Era",
    fill = "Era"
  ) +
  theme_minimal()

ggplotly(p_hard_hit, tooltip = "text", height = 420)
```

Average hard-hit rates rose from **30.5%** in the pre-Statcast era to **38.5%** in the Statcast era—a **26.2% relative increase** (an 8.0 percentage-point gain, climbing to 42.1% by 2025). Note that part of this shift coincides with the change in data measurement (scout observation vs. automated radar tracking).

