---
title: "Baseball Analysis"
execute:
  echo: false
  warning: false
---

```{r}
#| message: false
library(tidyverse)
library(plotly)

baseball_leagues <- read_csv("data/baseball_leagues.csv")

# Create cumulative data so lines draw forward across years as animation plays
create_anim_df <- \(df) {
  map_dfr(unique(df$Year), \(yr) {
    df |>
      filter(Year <= yr) |>
      mutate(Frame = yr)
  })
}

# Helper to generate spaghetti lines from 1,000 posterior simulation draws
create_spaghetti_df <- \(df, metric, n_sims = 1000, n_draws = 50) {
  fit <- lm(reformulate("Year", response = metric), data = df)
  set.seed(42)
  coefs <- MASS::mvrnorm(n_sims, mu = coef(fit), Sigma = vcov(fit))
  years <- seq(2005, 2025, by = 1)
  
  map_dfr(seq_len(n_draws), \(i) {
    tibble(
      draw = i,
      Year = years,
      val = coefs[i, 1] + coefs[i, 2] * years
    )
  }) |>
    group_by(draw) |>
    group_modify(\(data, keys) add_row(data, Year = NA, val = NA)) |>
    ungroup()
}

al_data <- baseball_leagues |> filter(League == "American League")
nl_data <- baseball_leagues |> filter(League == "National League")

anim_all <- create_anim_df(baseball_leagues)
anim_al  <- create_anim_df(al_data)
anim_nl  <- create_anim_df(nl_data)

# Spaghetti draws for SO/9
spag_al_so9 <- create_spaghetti_df(al_data, "SO9")
spag_nl_so9 <- create_spaghetti_df(nl_data, "SO9")

# Statcast Hard-Hit Rate data and simulation draws
hard_hit_obs <- read_csv("data/baseball.csv") |>
  filter(!is.na(Hard_Hit_pct)) |>
  select(Year, Hard_Hit_pct)

fit_hh <- lm(Hard_Hit_pct ~ Year, data = hard_hit_obs)

set.seed(42)
coefs_hh <- MASS::mvrnorm(1000, mu = coef(fit_hh), Sigma = vcov(fit_hh))
years_hh <- seq(2005, 2025, by = 1)

spag_hard_hit <- map_dfr(1:100, \(i) {
  tibble(
    draw = i,
    Year = years_hh,
    Hard_Hit_pct = coefs_hh[i, 1] + coefs_hh[i, 2] * years_hh
  )
})

league_colors <- c(
  "American League" = "#1f77b4",
  "National League" = "#d62728",
  "MLB Average"     = "#7f7f7f"
)

# Marker for the start of the Statcast era (2015)
statcast_shape <- list(
  type = "line",
  x0 = 2015,
  x1 = 2015,
  y0 = 0,
  y1 = 1,
  yref = "paper",
  line = list(color = "darkred", dash = "dash", width = 1.5)
)

statcast_annotation <- list(
  x = 2015.2,
  y = 0.95,
  yref = "paper",
  text = "2015: Statcast Era Begins",
  showarrow = FALSE,
  xanchor = "left",
  font = list(color = "darkred", size = 11)
)
```

## Strikeout Rise Over Time

Press **Play** or scrub the slider to animate strikeout rates from 2005 to 2025. Background spaghetti lines display sample trajectories from 1,000 posterior simulation draws to visualize model uncertainty. Switch tabs to compare leagues.

::: {.panel-tabset}
### Both Leagues

```{r}
#| label: plot-so9-both
plot_ly() |>
  add_trace(
    data = spag_al_so9,
    x = ~Year,
    y = ~val,
    type = "scatter",
    mode = "lines",
    line = list(color = "rgba(31, 119, 180, 0.18)", width = 1),
    hoverinfo = "none",
    showlegend = FALSE
  ) |>
  add_trace(
    data = spag_nl_so9,
    x = ~Year,
    y = ~val,
    type = "scatter",
    mode = "lines",
    line = list(color = "rgba(214, 39, 40, 0.18)", width = 1),
    hoverinfo = "none",
    showlegend = FALSE
  ) |>
  add_trace(
    data = anim_all,
    x = ~Year,
    y = ~SO9,
    color = ~League,
    colors = league_colors,
    frame = ~Frame,
    type = "scatter",
    mode = "lines+markers"
  ) |>
  layout(
    xaxis = list(range = c(2004, 2026), title = "Year"),
    yaxis = list(range = c(5.5, 9.5), title = "Strikeouts per 9 Innings (SO/9)"),
    shapes = list(statcast_shape),
    annotations = list(statcast_annotation)
  ) |>
  animation_opts(frame = 400, transition = 200, redraw = FALSE)
```

### American League

```{r}
#| label: plot-so9-al
plot_ly() |>
  add_trace(
    data = spag_al_so9,
    x = ~Year,
    y = ~val,
    type = "scatter",
    mode = "lines",
    line = list(color = "rgba(31, 119, 180, 0.22)", width = 1),
    hoverinfo = "none",
    showlegend = FALSE
  ) |>
  add_trace(
    data = anim_al,
    x = ~Year,
    y = ~SO9,
    color = ~League,
    colors = league_colors,
    frame = ~Frame,
    type = "scatter",
    mode = "lines+markers"
  ) |>
  layout(
    xaxis = list(range = c(2004, 2026), title = "Year"),
    yaxis = list(range = c(5.5, 9.5), title = "Strikeouts per 9 Innings (SO/9)"),
    shapes = list(statcast_shape),
    annotations = list(statcast_annotation)
  ) |>
  animation_opts(frame = 400, transition = 200, redraw = FALSE)
```

### National League

```{r}
#| label: plot-so9-nl
plot_ly() |>
  add_trace(
    data = spag_nl_so9,
    x = ~Year,
    y = ~val,
    type = "scatter",
    mode = "lines",
    line = list(color = "rgba(214, 39, 40, 0.22)", width = 1),
    hoverinfo = "none",
    showlegend = FALSE
  ) |>
  add_trace(
    data = anim_nl,
    x = ~Year,
    y = ~SO9,
    color = ~League,
    colors = league_colors,
    frame = ~Frame,
    type = "scatter",
    mode = "lines+markers"
  ) |>
  layout(
    xaxis = list(range = c(2004, 2026), title = "Year"),
    yaxis = list(range = c(5.5, 9.5), title = "Strikeouts per 9 Innings (SO/9)"),
    shapes = list(statcast_shape),
    annotations = list(statcast_annotation)
  ) |>
  animation_opts(frame = 400, transition = 200, redraw = FALSE)
```
:::

## Hard Hit Rate Over Time

To examine the progression across both the pre-tracking and Statcast eras, we model hard-hit rates from 2005 to 2025 using 1,000 posterior simulation draws.

The **spaghetti plot** below displays 100 sample trajectories (the thin, translucent lines) from 2005 to 2025, highlighting the shift at 2015 when Statcast tracking was officially introduced alongside observed data points (red).

```{r}
#| label: plot-hard-hit-spaghetti
ggplot() +
  geom_line(
    data = spag_hard_hit,
    aes(x = Year, y = Hard_Hit_pct, group = draw),
    alpha = 0.15,
    color = "steelblue"
  ) +
  geom_line(
    data = tibble(Year = years_hh, Hard_Hit_pct = coef(fit_hh)[1] + coef(fit_hh)[2] * years_hh),
    aes(x = Year, y = Hard_Hit_pct),
    color = "darkblue",
    linewidth = 1.2
  ) +
  geom_point(
    data = hard_hit_obs,
    aes(x = Year, y = Hard_Hit_pct),
    color = "darkred",
    size = 3.5
  ) +
  geom_vline(xintercept = 2015, linetype = "dashed", color = "darkred") +
  annotate(
    "text",
    x = 2015.3,
    y = 28,
    label = "2015: Statcast Era Begins",
    color = "darkred",
    hjust = 0
  ) +
  scale_x_continuous(breaks = seq(2005, 2025, by = 5)) +
  labs(
    title = "Hard Hit Rate Over Time (2005–2025)",
    subtitle = "Spaghetti plot of 100 sample lines from 1,000 posterior simulation draws",
    x = "Year",
    y = "Hard Hit %",
    caption = "Red points are observed Statcast measurements. Translucent lines display sample trajectories."
  ) +
  theme_minimal()
```
