---
title: "Baseball Analysis"
execute:
  echo: false
  warning: false
---

```{r}
#| message: false
library(tidyverse)
# Statcast & pre-Statcast Hard-Hit Rate data (player-seasons)
hitters_hard_hit <- read_csv("data/all_hitters_hard_hit.csv")

yearly_avg_hh <- hitters_hard_hit |>
  group_by(year, era) |>
  summarize(avg_hard_hit = mean(hard_hit_pct, na.rm = TRUE), .groups = "drop")

# Pitcher seasons data (K/9)
pitcher_seasons <- read_csv("data/pitcher_seasons.csv")

yearly_pitching <- pitcher_seasons |>
  group_by(yearID, era) |>
  summarize(avg_so9 = mean(so9), .groups = "drop")
```

## Strikeout Rates Over Time

Each point represents an individual pitcher-season (minimum 30 IP), displaying real pitcher-level variation across thousands of observations. The solid line and points trace the annual league average K/9 across the pre-Statcast and post-Statcast eras.

```{r}
#| label: plot-k9-pitcher-seasons
ggplot() +
  geom_jitter(
    data = pitcher_seasons,
    aes(x = yearID, y = so9, color = era),
    alpha = 0.12,
    width = 0.25,
    size = 1
  ) +
  geom_line(
    data = yearly_pitching,
    aes(x = yearID, y = avg_so9),
    color = "black",
    linewidth = 1.2
  ) +
  geom_point(
    data = yearly_pitching,
    aes(x = yearID, y = avg_so9, fill = era),
    shape = 21,
    size = 3,
    color = "black"
  ) +
  geom_vline(xintercept = 2014.5, linetype = "dashed", color = "darkred", linewidth = 1) +
  annotate(
    "text",
    x = 2014.8,
    y = 16,
    label = "Statcast Era Begins (2015)",
    color = "darkred",
    hjust = 0,
    size = 3.5
  ) +
  scale_x_continuous(breaks = seq(2005, 2025, by = 5)) +
  scale_color_manual(values = c("Pre-Statcast (2005–2014)" = "#377eb8", "Post-Statcast (2015–2025)" = "#e41a1c")) +
  scale_fill_manual(values = c("Pre-Statcast (2005–2014)" = "#377eb8", "Post-Statcast (2015–2025)" = "#e41a1c")) +
  labs(
    title = "Strikeouts per 9 Innings (K/9) by Pitcher-Season (2005–2025)",
    subtitle = "Comparing individual pitcher-seasons (points) and annual averages (line) across eras",
    x = "Year",
    y = "Average K/9",
    color = "Era",
    fill = "Era",
    caption = "Source: Lahman Baseball Database (min. 30 IP). Line indicates annual league average."
  ) +
  theme_minimal()
```

## Hard Hit Rate Over Time

Each point represents a qualifying hitter's individual hard-hit rate for that season, showing real individual-level variation across 4,251 player-seasons. The solid line and points trace the annual league average across the pre-Statcast era (BIS video-scouted Hard%) and the Statcast era (radar-measured exit velocity ≥ 95 mph).

```{r}
#| label: plot-hard-hit-player-seasons
ggplot() +
  geom_jitter(
    data = hitters_hard_hit,
    aes(x = year, y = hard_hit_pct, color = era),
    alpha = 0.15,
    width = 0.25,
    size = 1
  ) +
  geom_line(
    data = yearly_avg_hh,
    aes(x = year, y = avg_hard_hit, group = era),
    color = "black",
    linewidth = 1.2
  ) +
  geom_point(
    data = yearly_avg_hh,
    aes(x = year, y = avg_hard_hit, fill = era),
    shape = 21,
    size = 3,
    color = "black"
  ) +
  geom_vline(xintercept = 2014.5, linetype = "dashed", color = "darkred", linewidth = 1) +
  annotate(
    "text",
    x = 2014.8,
    y = 65,
    label = "Statcast Era Begins (2015)",
    color = "darkred",
    hjust = 0,
    size = 3.5
  ) +
  scale_x_continuous(breaks = seq(2005, 2025, by = 5)) +
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  scale_color_manual(values = c("Pre-Statcast (2005–2014)" = "#377eb8", "Post-Statcast (2015–2025)" = "#e41a1c")) +
  scale_fill_manual(values = c("Pre-Statcast (2005–2014)" = "#377eb8", "Post-Statcast (2015–2025)" = "#e41a1c")) +
  labs(
    title = "Hard-Hit Rate by Player-Season (2005–2025)",
    subtitle = "Comparing pre-Statcast (BIS scouted) vs. Statcast era (radar-measured ≥ 95 mph)",
    x = "Year",
    y = "Average Hard-Hit %",
    color = "Era",
    fill = "Era",
    caption = "Sources: FanGraphs / BIS (2005–2014) and Baseball Savant Statcast (2015–2025). Line indicates annual league average."
  ) +
  theme_minimal()
```
