---
title: "Interactive Comparison: Strikeouts vs. Hard-Hit Rate"
execute:
  echo: false
  warning: false
---

```{r}
#| message: false
library(tidyverse)
library(plotly)

# Load player-season datasets
hitters_hard_hit <- read_csv("data/all_hitters_hard_hit.csv")
trout_zone <- read_csv("data/trout_zone_whiff.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 MVP)", "Post-Statcast (2019 MVP)"))
trout_swings <- read_csv("data/trout_swings.csv") |>
  mutate(era = fct_relevel(era, "Pre-Statcast (2014 MVP)", "Post-Statcast (2019 MVP)"))
pitcher_seasons <- read_csv("data/pitcher_seasons.csv") |>
  left_join(
    Lahman::People |> select(playerID, nameFirst, nameLast) |> mutate(name = paste(nameFirst, nameLast)),
    by = "playerID"
  )

# Add horizontal jitter for individual points so they don't overlap completely
set.seed(42)
hitters_jittered <- hitters_hard_hit |>
  mutate(year_jitter = year + runif(n(), -0.25, 0.25))

pitchers_jittered <- pitcher_seasons |>
  mutate(year_jitter = yearID + runif(n(), -0.25, 0.25))

# Calculate annual league averages
yearly_hh <- hitters_hard_hit |>
  group_by(year) |>
  summarize(avg_hard_hit = round(mean(hard_hit_pct, na.rm = TRUE), 1))

yearly_k9 <- pitcher_seasons |>
  group_by(yearID) |>
  summarize(avg_so9 = round(mean(so9), 2))
```

## Interactive Era Comparison (2005–2025)

Use the filter buttons at the top to toggle between individual player points and clean league-average trends.

```{r}
#| label: plot-interactive-comparison
plot_ly(height = 480) |>
  # Hitter player-seasons (left axis)
  add_markers(
    data = hitters_jittered,
    x = ~year_jitter,
    y = ~hard_hit_pct,
    name = "Hitter Seasons",
    marker = list(color = "#e41a1c", opacity = 0.2, size = 3.5),
    text = ~paste0("<b>", name, "</b> (", year, ")<br>Hard-Hit Rate: ", hard_hit_pct, "%"),
    hoverinfo = "text"
  ) |>
  # Pitcher player-seasons (right axis)
  add_markers(
    data = pitchers_jittered,
    x = ~year_jitter,
    y = ~so9,
    yaxis = "y2",
    name = "Pitcher Seasons",
    marker = list(color = "#377eb8", opacity = 0.2, size = 3.5),
    text = ~paste0("<b>", name, "</b> (", yearID, ")<br>K/9: ", round(so9, 2)),
    hoverinfo = "text"
  ) |>
  # Annual league average lines
  add_lines(
    data = yearly_hh,
    x = ~year,
    y = ~avg_hard_hit,
    name = "Avg Hard-Hit %",
    line = list(color = "#b30000", width = 3),
    text = ~paste0("<b>", year, " League Avg</b><br>Hard-Hit Rate: ", avg_hard_hit, "%"),
    hoverinfo = "text"
  ) |>
  add_lines(
    data = yearly_k9,
    x = ~yearID,
    y = ~avg_so9,
    yaxis = "y2",
    name = "Avg K/9",
    line = list(color = "#08519c", width = 3),
    text = ~paste0("<b>", yearID, " League Avg</b><br>K/9: ", avg_so9),
    hoverinfo = "text"
  ) |>
  layout(
    title = list(
      text = "<b>Player-Season Hard-Hit Rate vs. Strikeout Rate</b> (2005–2025)",
      x = 0.5,
      xanchor = "center",
      y = 0.98,
      font = list(size = 15)
    ),
    margin = list(t = 80, b = 60, l = 60, r = 60),
    xaxis = list(title = "Year", dtick = 5, gridcolor = "#f0f0f0"),
    yaxis = list(
      title = list(text = "Hard-Hit Rate (%)", font = list(color = "#b30000", size = 12)),
      ticksuffix = "%",
      gridcolor = "#f0f0f0"
    ),
    yaxis2 = list(
      title = list(text = "Strikeouts per 9 Innings (K/9)", font = list(color = "#08519c", size = 12)),
      overlaying = "y",
      side = "right",
      showgrid = FALSE
    ),
    hoverlabel = list(bgcolor = "#2c3e50", font = list(color = "white")),
    # Interactive view switcher buttons
    updatemenus = list(
      list(
        type = "buttons",
        direction = "right",
        x = 0.5,
        xanchor = "center",
        y = 1.10,
        buttons = list(
          list(method = "restyle", args = list("visible", list(TRUE, TRUE, TRUE, TRUE)), label = "Show All"),
          list(method = "restyle", args = list("visible", list(FALSE, FALSE, TRUE, TRUE)), label = "Averages Only"),
          list(method = "restyle", args = list("visible", list(TRUE, FALSE, TRUE, FALSE)), label = "Hitters Only"),
          list(method = "restyle", args = list("visible", list(FALSE, TRUE, FALSE, TRUE)), label = "Pitchers Only")
        )
      )
    ),
    # Shaded backgrounds for pre-Statcast and post-Statcast eras
    shapes = list(
      list(
        type = "rect",
        x0 = 2004.5,
        x1 = 2014.5,
        y0 = 0,
        y1 = 1,
        yref = "paper",
        fillcolor = "rgba(55, 126, 184, 0.06)",
        line = list(width = 0),
        layer = "below"
      ),
      list(
        type = "rect",
        x0 = 2014.5,
        x1 = 2025.5,
        y0 = 0,
        y1 = 1,
        yref = "paper",
        fillcolor = "rgba(228, 26, 28, 0.06)",
        line = list(width = 0),
        layer = "below"
      ),
      list(
        type = "line",
        x0 = 2014.5,
        x1 = 2014.5,
        y0 = 0,
        y1 = 1,
        yref = "paper",
        line = list(dash = "dash", color = "#777777", width = 1.5)
      )
    ),
    annotations = list(
      list(
        x = 2009.5,
        y = 0.98,
        yref = "paper",
        text = "<b>Pre-Statcast Era</b>",
        showarrow = FALSE,
        font = list(color = "#377eb8", size = 12)
      ),
      list(
        x = 2020,
        y = 0.98,
        yref = "paper",
        text = "<b>Statcast Era (2015+)</b>",
        showarrow = FALSE,
        font = list(color = "#e41a1c", size = 12)
      )
    ),
    legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.16)
  ) |>
  config(responsive = TRUE)
```

By displaying both sets of player-seasons together, we can observe how the entire distribution of player performances shifted upward along with the introduction of Statcast in 2015.

## Game Evolution: Strikeout vs. Hard-Hit Trajectory

Instead of tracking both metrics on a shared timeline, this **connected scatter plot (trajectory plot)** graphs them directly against each other. Tracing the path chronologically from 2005 (bottom left) to 2025 (top right) visualizes how Major League Baseball migrated diagonally upward—moving toward higher strikeout rates and higher exit velocities at the same time.

```{r}
#| label: plot-game-trajectory
combined_trajectory <- inner_join(yearly_hh, yearly_k9, by = c("year" = "yearID")) |>
  mutate(era = if_else(year < 2015, "Pre-Statcast (2005–2014)", "Post-Statcast (2015–2025)"))

plot_ly(data = combined_trajectory, x = ~avg_so9, y = ~avg_hard_hit, height = 420) |>
  # Chronological path connecting seasons
  add_paths(
    line = list(color = "#888888", width = 2, dash = "solid"),
    showlegend = FALSE,
    hoverinfo = "none"
  ) |>
  # Season marker points colored by era
  add_markers(
    color = ~era,
    colors = c("Pre-Statcast (2005–2014)" = "#377eb8", "Post-Statcast (2015–2025)" = "#e41a1c"),
    marker = list(size = 9, line = list(color = "black", width = 1)),
    text = ~paste0("<b>Season: ", year, "</b><br>Era: ", era, "<br>Strikeouts per 9: ", avg_so9, "<br>Hard-Hit Rate: ", avg_hard_hit, "%"),
    hoverinfo = "text"
  ) |>
  # Year labels for key seasons
  add_text(
    text = ~year,
    textposition = "top center",
    showlegend = FALSE,
    textfont = list(size = 9, color = "#333333")
  ) |>
  layout(
    title = list(
      text = "<b>The Shift in Modern Baseball (2005–2025)</b><br><sup>League Average K/9 vs. Hard-Hit Rate Trajectory</sup>",
      x = 0.5,
      xanchor = "center",
      font = list(size = 15)
    ),
    margin = list(t = 70, b = 60, l = 60, r = 50),
    xaxis = list(
      title = "League Average Strikeouts per 9 Innings (K/9)",
      dtick = 0.5,
      gridcolor = "#f0f0f0"
    ),
    yaxis = list(
      title = "League Average Hard-Hit Rate (%)",
      ticksuffix = "%",
      gridcolor = "#f0f0f0"
    ),
    annotations = list(
      list(
        x = 6.55, y = 30.2,
        text = "<b>2005 (Start)</b>",
        showarrow = TRUE,
        arrowhead = 2,
        ax = -45, ay = 25
      ),
      list(
        x = 8.61, y = 42.1,
        text = "<b>2025 (Latest)</b>",
        showarrow = TRUE,
        arrowhead = 2,
        ax = 45, ay = -25
      )
    ),
    legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.18)
  ) |>
  config(responsive = TRUE)
```

## Strike Zone Whiff Rate: Mike Trout (Pre- vs. Post-Statcast MVP)

The evolution between Mike Trout's 2014 and 2019 strike zones illustrates how tracking technology created a data-driven arms race between hitters and pitchers. In the early Statcast era, teams utilized high-speed optical cameras and radar to measure pitch spin rates and induced vertical break, discovering that elevated fastballs were uniquely effective at generating whiffs against modern uppercut swing paths. Concurrently, hitters used granular pitch-tracking feedback and video analytics to eliminate swing inefficiencies, allowing Trout to virtually master the middle of the plate—his whiff rate in the heart of the zone dropped from **12.2% in 2014 down to just 4.9% in 2019**. In response, pitchers adjusted their scouting heat maps away from the center and attacked the upper quadrant with high-spin velocity, elevating his whiff rate at the top of the zone to **23.0%**. This shift demonstrates the broader impact of Statcast: as tracking data helped elite hitters dominate hittable pitches, pitchers adapted by targeting the extreme vertical boundaries of the strike zone.

```{r}
#| label: plot-trout-strikezone
#| fig-width: 8.5
#| fig-height: 4.8
ggplot() +
  # Semi-transparent zone tiles colored by whiff rate
  geom_tile(
    data = trout_zone,
    aes(x = factor(x), y = factor(y), fill = whiff_pct),
    alpha = 0.35,
    color = "grey40",
    linewidth = 1.2
  ) +
  # Individual swing dots jittered inside their respective zone
  geom_jitter(
    data = trout_swings,
    aes(x = factor(x), y = factor(y), color = outcome),
    width = 0.35,
    height = 0.35,
    alpha = 0.65,
    size = 1.8
  ) +
  # Zone whiff % labels
  geom_label(
    data = trout_zone,
    aes(x = factor(x), y = factor(y), label = paste0(round(whiff_pct, 1), "%")),
    alpha = 0.75,
    fontface = "bold",
    size = 3.2,
    color = "black"
  ) +
  scale_fill_gradient(low = "#4575b4", high = "#d73027", name = "Zone Whiff %") +
  scale_color_manual(values = c("Contact / Foul" = "#7f7f7f", "Whiff (Miss)" = "#d73027"), name = "Swing Outcome") +
  scale_x_discrete(labels = c("Inside", "Middle", "Away")) +
  scale_y_discrete(labels = c("Low", "Middle", "High")) +
  facet_wrap(~era) +
  coord_fixed() +
  labs(
    title = "Mike Trout Strike Zone: Individual Swings & Whiff Rates",
    subtitle = "Points represent individual swings (grey = contact/foul, red = whiff) over zone whiff %",
    x = "Horizontal Location (to RHB)",
    y = "Vertical Location"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 13)
  )
```

Between 2014 and 2019, Trout virtually eliminated whiffs in the heart of the zone—dropping from a **12.2% whiff rate** down to **4.9%**. Concurrently, pitchers shifted to attacking him with high fastballs, where his swing-and-miss rate increased to **23%** in the upper-inside quadrant.


